HMMotion: Using tracking data to predict coverage strategies in American Football

### 2025-01-06 submission.Rmd
.

<!--We analyze tracking data from the first nine weeks of the NFL 2022 season, provided by the NFL Big Data Bowl 2024. Moreover, we use play-by-play data and the corresponding data from PFF that assigned the categories \textit{man}, \textit{zone} and \textit{others} representing the different schemes to each play. As it is not properly described what \textit{others} means, we omit every play that is associated with this value. We also omit plays with more than five offensive linemen and with two quarterbacks. Since we are specifically interested in analyzing pre-snap player movements, we concentrate on plays that contain any pre-snap motion. Then, we end up with $3985$ offensive plays in total, from which the defense played $2973$ in zone and $1012$ in man coverage. -->

@@

To accurately forecast the defensive scheme (man- or zone defense) for every play, we create various features derived from the tracking data. In particular, we conducted the following feature engineering steps: First, using all 11 players on each side, we compute the area spanned by the convex hull of a team as well as the largest $y$ distance (i.e.\ the width of the hull) and the largest $x$ distance (i.e. the length of the hull). Then, we select the five most relevant players on each side of the field. For offense, we omit the offensive line and the QB, thus coming up with five players that are a composition of running backs, tight ends and wide receivers. In contrast, for defense, disregarding defense liners (NT, DT, DE), we select the five defenders that were the closest to the five offensive players, corresponding to a weighted euclidean distance, putting much more emphasis on the y-axis. From these 10 players, we derive features related to their (standardized) position, distance and orientation.
<!--
For all 11 players on each side of the field we compute features related to the convex hull of the positions of the players. In particular, for defense and offense, we compute the area spanned by the convex hull of all player such as well as the largest $y$ distance (i.e.\ the width of the hull) and the largest $x$ distance (i.e. the length of the hull). In addition, we select the five most relevant players on each side of the field. For offense, we omit the offensive line and the QB, thus coming up with five players that are a composition of running backs, tight ends and wide receivers. In contrast, for defense, we omit nose tackles, defensive tackles and defensive ends, and select the five defenders that were the closest to the five offensive players, corresponding to a weighted euclidean distance, putting much more emphasis on the y-axis. 
For these 10 players, we extract features related to their (standardized) position, distance and orientation.
<!--
Finally, we use their standardized $x$ and $y$ coordinates as features and order defensive and offensive players according to their $y$ coordinates. Additionally, for each of the relevant defenders, we compute distances to the football and their orientation with respect to the quarterback. 
-->
Additionally, we extract relevant information from the play-by-play data, such as quarter, down, yards to go, home and away score and the remaining game time in the current half (in seconds). See the Appendix for a more detailed description of the features.

@@

<!--
`r colorize("der erste Teil ist doch eigentlich eine Wiederholung des Feature Engeneerings?","red")`
In general, we have a limited dataset available (only 3985 plays) and therefore need to balance the complexity of the model, thus aiming to control the number of features. Given the small dataset, we focus on 32 previously described basic features used in all 3 models: 6 convex hull related features (3 for offense and defense, respectively), 20 player positions features (10 standardized $x$ and $y$ coordinates, 5 for offense, 5 for defense), and 6 play-by-play features. In the Appendix, we provide a discussion on the choice of features. <!--we provide more details and analyses using more features. -->

@@

First of all, we need a suitable basic model class for predicting man or zone coverage and opt for the following two: First, we fit a `glmnet` (elastic net) model, which performs implicit feature selection and is able to handle multicollinearity. Second, we use an `xgboost` model, which is able to capture non-linear effects (and interactions) and also handles multicollinearity. However, it necessitates careful hyperparameter tuning and generally performs better on bigger data sets. For all of the models, we use 10-fold cross validation on a suitable hyperparameter grid. `r colorize("Ich finde der Teil ist besser in der Einleitung zur Analyse, da wir die zwei Modellklassen ja für alle 3 Modelle verwenden. Daher auch nochmal die Beschreibung der Features oben, weil die 32 features sind einfach für alle Modell gleich.","red")`

We fit the models with the previously described basic features. All of these features are derived at the time of line-set, which is why these models do not use any pre-snap motion information. These very basic models serve as baseline models that allows to measure the effect of pre-snap motion (features) in the following.

@@

In a second step, we extend our basic pre-motion model with naive post-motion features. To keep the complexity manageable, we derive only 6 additional post-motion features: for each team (offense and defense), we infer the maximum $y$-distance, the maximum $x$-distance and the total distance traveled by both teams until the snap.

@@

Specifically, we use results derived from an HMM fitted to the pre-snap motion data as further features in the previously described models. To achieve this, we model the movements of the five defensive players during pre-snap motion by an HMM (see the Appendix for an in-depth description). In particular, we assume that each defender's $y$-coordinate at each time point $t$ is a realization from a Gaussian distribution with mean aligned to the $y$-coordinate of the offensive player they are guarding and an estimated standard deviation. HMMs are particularly well-suited for this task since the guarding assignments are not directly observed but need to be inferred from the defenders' responses to the offenders' movements.
Hence, the model naturally treats this information as a latent state variable, making each observation a realization from a mixture of Gaussian distributions (see Franks et al. 2015 for a similar approach in basketball).
The ultimate goal is to use the fitted model for *state-decoding*, i.e., inferring information on the guarding assignment based on the observations made. HMMs excel in this context as they leverage not only the defenders' $y$-coordiate as well as the $y$-coordiate of all offenders potentially guarded at the current time point, but also incorporate probabilistic information based on the previous and subsequent observations. State-decoding can either provide us with fully probabilistic information, i.e., the probability of each defender guarding each offensive player at each time point, or a single most likely guarding assignment for each defender at each time point. We will use the both to enhance our post-motion models and use the latter for visualization purposes.
For details on model fitting and state decoding, also see the Appendix.

@@

To incorporate pre-snap information, we re-train the post-motion model to predict whether the defense employs a man- or zone-coverage scheme, now integrating results from the HMM analysis as additional features. A caveat of the aforementioned decoded state probabilities is that they form a multi-dimensional time series, complicating their direct use as features in the previously described models. This challenge can however be remedied by employing suitable summary statistics. To achieve this, we first calculate for each defender the most likely offensive player $k = 1, \ldots, 5$ to be guarded in each time point $t = 1, \ldots, T_n$, where $T_n$ is the length of the $n$-th play, and count the number of state switches, i.e. the number of times for which a defender's most likely guarded offensive players differs between two consecutive time points $t$ and $t+1$. From this, we can calculate some simple summary statistics, specifically 1) the sum of state switches, 2) the average number of state switches, and 3) the number of defenders that switch offensive players during a play. Additionally, we use the aforementioned decoded state probabilities to calculate a more elaborated statistic, the mean entropy across defenders $j = 1, \ldots, 5$ in each offensive play $n = 1,\ldots, N$:

@@

<div style="text-align: center;">
  ![](ACCs_bdb_main2.png)
  ![](AUCs_bdb_main2.png)
</div>

@@

![](differences_teams.png)

@@

          columns = list(
            url = colDef(
              name = "Team",
              maxWidth = 70,
              align = "center",
              cell = embed_img(height = 25, width = 40)
            ),
            n = colDef(
              name = "# Motions",
              format = colFormat(digits = 0),
              style = color_scales(team_pals, color_ref = "col1"),
              maxWidth = 90
            ),
            n_improv = colDef(
              name = "# Plays with improved probability of predicting correct defensive scheme in post motion model",
              format = colFormat(digits = 0),
              style = color_scales(team_pals, color_ref = "col2")
            ),
            prop_improv = colDef(
              name = "Percentation of plays with improved probability of predicting correct defensive scheme in post motion model",
              format = colFormat(digits = 2),
              style = color_scales(team_pals, color_ref = "col3")
            )
          #   avg_per_tackle2 = colDef(
          #     name = "Avg PEP[2]",
          #     format = colFormat(digits = 2),
          #     style = color_scales(best_tacklers_u, color_ref = "col4"),
          #     maxWidth = 80
          #   )
          )

@@

While we were clearly able to show that pre-snap motion facilitates the defensive coverage scheme detection, a limitation of our approach lies in the imperfect prediction accuracy of the pre-motion model, primarily due to insufficient hyperparameter tuning and the relatively small number of plays involving motion. `r colorize("Ich würde nicht mit limitations anfangen sondern mit dem was wir gezeigt haben. Außerdem sollte man das insufficient hyperparameter tuning weglassen und nur auf den kleinen Datensatz fokussieren. Wir tunen gut genug und es liegt auch am Datensatz, dass wir nicht mehr machen.","red")` However, the primary focus of this project was on pre-snap motion, particularly on how to effectively translate this information into the hidden Markov model. Importantly, our HMM approach is modular and can be seamlessly replaced by another model, such as the NFL Next Gen Stats model, in the case of a richer data set. Combining both of these worlds could fully leverage the insights provided by the present data.

`r colorize("Vorschlag Robert (mit Teilen von oben und chatGPT Hilfe von unten):","red")`

@@

Comparing predictions from pre-motion and post-motion models provides insights into an offense's ability to use motion effectively to identify defensive coverage schemes. Similarly, it allows for evaluating a defense's skill in disguising their coverage and confusing the offense during pre-snap motion.

While our pre- and post-motion models are limited due to small sample size available, our HMM approach is modular and therefore our models can be replaced by more another model in the case of a richer data set. Combining both of these worlds could fully leverage the insights provided by the present data.

`r colorize("Von chatGPT:","red")`

This project explored the use of hidden Markov models (HMMs) to enhance the prediction of defensive schemes in football by incorporating pre-snap player movements. Models augmented with HMM-derived features demonstrated improved predictive performance compared to those relying solely on naive motion data. The HMM effectively captured latent defensive behaviors, offering deeper insights into the defenders' actions.

The HMM features allowed for more effective inference of guarding assignments, improving the distinction between man and zone coverage. This improvement underscores the HMM's capacity to model the dynamic and often subtle nature of defensive strategies, capturing patterns that simple motion features cannot.

Looking ahead, the potential of HMMs goes beyond predicting man or zone defense. They could be applied to decode complex guarding assignments and analyze defensive adjustments in real-time, providing a powerful tool for deeper strategic insights in sports analytics.

@@

Prior to more involved feature engeneering steps, we transform the coordinate system by redefining the x-variable as the x-distance to the endzone (such that all play directions are from right to left and the relevant endzone is at zero), and changing the direction variable, such that zero degrees represents heading straight towards the corresponding endzone. As mentioned in the main text, we use player features from 5 defensive and 5 offensive players. First, we standardized their $x$- and $y$-coordinates with respect to the football and ordered the players according to their $y$-coordinates, i.e., the the first defender in our dataset is always the leftmost defensive player, while the first offensive player is always the rightmost one (offensive play direction is from left to right). Furthermore, for each player we compute distances to the football and their orientation with respect to the quarterback.

@@

As frequently mentioned, we face the problem of only having a small amount of relevant data available. In order to avoid overfitting problems, we therefore focus on a basic feature set of 32 variables for our main results: 6 convex hull related features (3 for offense and defense, respectively), 20 player positions features (10 standardized $x$ and 10 standardized $y$ coordinates, 5 for offense, 5 for defense), and 6 play-by-play features. However, using more thoroughly crafted features such as distances and orientation as described above, we can enlarge the feature set to 67 total variables: 30 distance variables (for each of the 10 relevant players the total distance, $x$-distances, and $y$-distances to the football), and 5 orientation variables (for each defender the orientation with respect to the QB).

In the following, we provide a comparison of these enlarged models and the smaller ones used for our main results. When comparing results of a binary outcome variable, accuracy can be a misleading metric, since it is dependent on the threshold selected for classification of the outcome. Usually, for probabilistic predictions as obtained by the two model classes considered, one uses the naive threshold of 0.5. However doing so is arbitrary and especially for imbalanced data, changing the threshold may change the results of the accuracy drastically. AUC (or area under the ROC curve) on the other hand, tries to evaluated the performance of a model over a suitable grid of thresholds, and thus avoids the subjective choice of threshold. Another popularly used metric is the logloss (or negative log-likelihood loss), which is often a preferred choice due to being a proper scoring rule. While using the logloss as evaluation metric is mathematically the best option, interpretation is not as intuitive. Specifically, whether a logloss of a model can be considered good, depends on the (im)balance of the classes (as well as the number of classes, which in a our binary case is only 2). Without going into further detail, we therefore focus on these two metrics for the evaluation of the models.

>>>>

@@

To accurately forecast the defensive scheme, we create various features derived from the tracking data. In particular, we conducted the following feature engineering steps: First, using all 11 players on each side, we compute the area spanned by the convex hull of a team as well as the largest $y$ distance (i.e.\ the width of the hull) and the largest $x$ distance (i.e. the length of the hull). Then, we select the five most relevant players on each side of the field. For offense, we omit the offensive line and the QB, thus coming up with five players that are a composition of running backs, tight ends and wide receivers. In contrast, for defense, disregarding defense liners (NT, DT, DE), we select the five defenders that were the closest to the five offensive players, corresponding to a weighted euclidean distance, putting much more emphasis on the y-axis. From these 10 players, we derive features related to their (standardized) position, distance and orientation.

Additionally, we extract relevant information from the play-by-play data, such as quarter, down, yards to go, home and away score and the remaining game time in the current half (in seconds). See the Appendix for a more detailed description of the features.

@@

First of all, we need a suitable basic model class for predicting man or zone coverage and opt for the following two: First, we fit a `glmnet` (elastic net) model, which performs implicit feature selection and is able to handle multicollinearity. Second, we use an `xgboost` model, which is able to capture non-linear effects (and interactions) and also handles multicollinearity. However, it necessitates careful hyperparameter tuning and generally performs better on bigger data sets. For all of the models, we use 10-fold cross validation on a suitable hyperparameter grid.

@@

At first, we fit the aforementioned models with the previously described basic pre-motion features as all of these features are derived at the time of line-set. These very basic models serve as baseline models that allows to measure the effect of pre-snap motion (features) in the following.

@@

In a second step, we extend our basic pre-motion model with naive post-motion features. To keep the complexity manageable, we derive only six additional post-motion features: for each team (offense and defense), we infer the maximum $y$-distance, the maximum $x$-distance and the total distance traveled by both teams until the snap.

@@

Specifically, we use results derived from an HMM fitted to the pre-snap motion tracking data as further features in the previously described models. To achieve this, we model the movements of the five defensive players during pre-snap motion by an HMM (see the Appendix for an in-depth description). In particular, we assume that each defender's $y$-coordinate at each time point $t$ is a realization from a Gaussian distribution with mean aligned to the $y$-coordinate of the offensive player they are guarding and an estimated standard deviation. HMMs are particularly well-suited for this task since the guarding assignments are not directly observed but need to be inferred from the defenders' responses to the offenders' movements.
Hence, the model naturally treats this information as a latent state variable, making each observation a realization from a mixture of Gaussian distributions (see Franks et al. 2015 for a similar approach in basketball).
The ultimate goal is to use the fitted model for *state-decoding*, i.e., inferring information on the guarding assignment based on the observations made. HMMs excel in this context as they leverage not only the defenders' $y$-coordiate and the $y$-coordiate of all offenders potentially guarded at the current time point, but also incorporate probabilistic information based on the previous and subsequent observations. State-decoding can either provide us with fully probabilistic information, i.e., the probability of each defender guarding each offensive player at each time point, or a single most likely guarding assignment for each defender at each time point. We will use the former to enhance our post-motion models and the latter for visualization purposes.
For details on model fitting and state decoding, we also refer to the Appendix.

@@

To incorporate pre-snap motion information, we re-train the post-motion model to predict whether the defense employs a man- or zone-coverage scheme, now integrating results from the HMM analysis as additional features. A caveat of the aforementioned decoded state probabilities is that they form a multi-dimensional time series, complicating their direct inclusion as features in the previously described models. This challenge can however be remedied by employing suitable summary statistics. To achieve this, we first calculate for each defender the most likely offensive player $k = 1, \ldots, 5$ to be guarded in each time point $t = 1, \ldots, T_n$, where $T_n$ is the length of the $n$-th play, and count the number of state switches, i.e. the number of times for which a defender's most likely guarded offensive players differs between two consecutive time points $t$ and $t+1$. From this, we can calculate some simple summary statistics, specifically 1) the sum of state switches, 2) the average number of state switches, and 3) the number of defenders that switch offensive players during a play. Additionally, we use the aforementioned decoded state probabilities to calculate a more elaborated statistic, the mean entropy across defenders $j = 1, \ldots, 5$ in each offensive play $n = 1,\ldots, N$:

@@

<div style="text-align: center;">
  ![](ACCs_bdb_main.png)
  ![](AUCs_bdb_main.png)
</div>

@@

![](differences_teams.png){width=50%}

@@

          columns = list(
            url = colDef(
              name = "Team",
              maxWidth = 70,
              align = "center",
              cell = embed_img(height = 25, width = 40)
            ),
            n = colDef(
              name = "# Motions",
              format = colFormat(digits = 0),
              style = color_scales(team_pals, color_ref = "col1"),
              maxWidth = 90
            ),
            n_improv = colDef(
              name = "# Plays with improved probability of predicting correct defensive scheme in HMM post-motion model",
              format = colFormat(digits = 0),
              style = color_scales(team_pals, color_ref = "col2")
            ),
            prop_improv = colDef(
              name = "Percentation of plays with improved probability of predicting correct defensive scheme in HMM post-motion model",
              format = colFormat(digits = 2),
              style = color_scales(team_pals, color_ref = "col3")
            )
          #   avg_per_tackle2 = colDef(
          #     name = "Avg PEP[2]",
          #     format = colFormat(digits = 2),
          #     style = color_scales(best_tacklers_u, color_ref = "col4"),
          #     maxWidth = 80
          #   )
          )

@@

@@

Comparing predictions from pre-motion and post-motion models provides insights into an offense's ability to use motion effectively to identify defensive coverage schemes. Similarly, future analysis could also relate to evaluating a defense's skill in disguising their coverage and confusing the offense during pre-snap motion.

While our pre- and post-motion models are limited due to small sample size available, our HMM approach is modular and therefore our models can be seamlessly replaced by other models, potentially more complex ones in the case of a richer data set. Combining both of these worlds could fully leverage the insights provided by the present data.

@@

Prior to more involved feature engeneering steps, we transform the coordinate system by redefining the x-variable as the x-distance to the endzone (such that all play directions are from right to left and the relevant endzone is at zero), and changing the direction variable, such that zero degrees represents heading straight towards the corresponding endzone. As mentioned in the main text, we use player features from 5 defensive and 5 offensive players. First, we standardized their $x$- and $y$-coordinates with respect to the football and ordered the players according to their $y$-coordinates, i.e., the first defender in our dataset is always the leftmost defensive player, while the first offensive player is always the rightmost one (offensive play direction is from left to right). Furthermore, for each player we compute distances to the football and their orientation with respect to the quarterback.

@@

As frequently mentioned, we face the problem of only having a small amount of relevant data available. To avoid overfitting problems, we therefore focus on a basic feature set of 32 variables for our main results: 6 convex hull related features (3 for offense and defense, respectively), 20 player positions features (10 standardized $x$ and 10 standardized $y$ coordinates, 5 for offense, 5 for defense), and 6 play-by-play features. However, using more thoroughly crafted features such as distances and orientation as described above, we can enlarge the feature set to 67 total variables: 30 distance variables (for each of the 10 relevant players the total distance, $x$-distances, and $y$-distances to the football), and 5 orientation variables (for each defender the orientation with respect to the QB).

In the following, we provide a comparison of these enlarged models and the smaller ones used for our main results. When comparing results of a binary outcome variable, accuracy can be a misleading metric, since it is dependent on the threshold selected for classification of the outcome. Usually, for probabilistic predictions as obtained by the two model classes considered, one uses the naive threshold of $0.5$. However doing so is arbitrary and especially for imbalanced data, changing the threshold may change the results of the accuracy drastically. AUC (or area under the ROC curve) on the other hand, tries to evaluated the performance of a model over a suitable grid of thresholds, and thus avoids the subjective choice of threshold. Another popularly used metric is the logloss (or negative log-likelihood loss), which is often a preferred choice due to being a proper scoring rule. While using the logloss as evaluation metric is mathematically the best option, interpretation is not as intuitive. Specifically, whether a logloss of a model can be considered good, depends on the (im)balance of the classes (as well as the number of classes, which in a our binary case is only 2). Without going into further detail, we therefore focus on these two metrics for the evaluation of the models.
